\LARGE \bf Experimental Validation of Reference Spreading by Robotic Manipulation of Unmodeled Objects

### 2023-01-07 PaperPreliminaryV2.tex
aha

% Setup
    \documentclass[a4paper, 10pt, conference]{ieeeconf}
    \IEEEoverridecommandlockouts                              % This command is only
                                                              % needed if you want to
                                                              % use the \thanks command
    \overrideIEEEmargins
    % See the \addtolength command later in the file to balance the column lengths
    % on the last page of the document

@@

    % The following packages can be found on http:\\www.ctan.org
    \usepackage{graphics} % for pdf, bitmapped graphics files
    \usepackage{epsfig} % for postscript graphics files
    \usepackage{mathptmx} % assumes new font selection scheme installed
    \usepackage{mathptmx} % assumes new font selection scheme installed
    \usepackage{amsmath} % assumes amsmath package installed
    \usepackage{amssymb}  % assumes amsmath package installed

    \title{\LARGE \bf
    Experimental Validation of Reference Spreading by Robotic Manipulation of Unmodeled Objects
    }

@@

    \subsection{Quadratic programing control} \label{sec:QP}
     \textcolor{red}{goal: motivate QP control, formulate carthesian impedance task, posture task, and list dynamics of motion and safety constraints }\\
     Controlling a robot for object manipulation involves, e.g., adressing redundancy and acounting for limits of the robot, while simultaneously applying a contact force and moving to a position. These objectives often clash with each other, meaning that they cannot always be perfectly satisfied. It is possible to distinguish between constraints -- control objectives that must always be satisfied -- and tasks -- objectives that should be performed as good as possible but not necesarily perfectly.

     Quadratic Programming (QP) relies on this philosophy. In a QP scheme, constraints and task errors are defined as a linear function of optimization parameters. Then it becomes possible to efficiently find the parameters that minmize the task errors while adhering to the constraints. Furthermore, priority between the tasks can be achieved by assigning weights. 
     In the context of a robot where the optimization parameters are, for example, joint accelerations $\ddot{q}$, the optimization problem can be formulated as

@@

     \textbf{Torque constraint:} To protect the robot, we can define a constraint that limits the control torque $\tau$ within a safe range, i.e.,
     \begin{equation}
       \tau_{min} \leq\tau \leq  \tau_{max} 
     \end{equation}
     This constraint must be formulated as a function of optimization parameter $\ddot q$ to be compatible with the QP scheme. The relation between $\ddot{q}$ and $\tau$ is given by the equations of motion
     \begin{equation}
        M\ddot{q}+h=\tau
     \end{equation}    
     with inertia matrix $M$ and vector of gravity, centrifugal, and Coriolis terms $h$. Subsequently, the torque constraint formulated as linear function of the optimization parameter becomes
    \begin{equation}
     \tau_{min} \leq M\ddot{q}+h \leq  \tau_{max}.       
     \end{equation}

     \textbf{Impedance task:} When manipulating objects, both the position and the contact force of the robot should be controlled. Impedance relates displacement to force, meaning that enforcing a desired impedance is a suitable control approach for object manipulation.

     Rather than enforcing a virtual impedance at each joint, it is preferable to apply a virtual impedance in Carthesian space, which is more intuitive when tuning task weights or when performing teleoperation. The impedance should then act on the carthesian pose of a body, i.e., its translational coordinates $p$ and rotation $R$. Velocity and accelerations are represented by $\omega$ and $\alpha$ for rotations, and by $v$ and $a$ for translations.

     An impedance controller aims to achieve the closed loop behaviour
         \begin{equation} \label{eq:impedance_desired}
     \Lambda  \begin{bmatrix} \alpha   \\ a  \end{bmatrix} + D \begin{bmatrix} \omega - {\omega_d} \\ v - v_d \end{bmatrix}  + K \begin{bmatrix} -e_{rot} \\p - {p_d}  \end{bmatrix} = f_d
     \end{equation}

    with Carthesian-space inertia, damping and stiffness matrices $\Lambda$, $D$ and $K$, and rotation tracking error $e_{rot}$ as defined in Appendix xxx. Subscript $d$ denotes the desired value, and $f$ represents the external wrench.

    Stiffness $K$ is typically chosen as a diagonal matrix. The damping matrix is determined following $D = 2(\Lambda K)^{\frac{1}{2}}$ which guarantees stable behavior when $K$ and $\Lambda$ are symmetric~\cite{albu-schafferCartesianImpedanceControl2003}. Furthermore, for the inertia matrix, two options are considered. Choosing a diagonal matrix $\Lambda$ decouples the accelerations w.r.t. to the position error, resulting in better position tracking in free motion. This approach was used in~\cite{vanoorschotDesignNumericalValidation2022}. In this work, however, the task-space inertia is set to match the joint-space inertia following $\Lambda^{-1} = JM^{-1}J^T$, decoupling the contact force w.r.t. the position error for better performance during contact.

    We can define the target task-space accelerations $\alpha_t$ and $a_t$ based on \eqref{eq:impedance_desired}:
    \begin{align} \label{eq:imptask1}
    \begin{bmatrix} \alpha_{t} \\ a_{t} \end{bmatrix} 
   =    \Lambda^{-1} \left ( D \begin{bmatrix} \omega_d - {\omega} \\ v_d - v \end{bmatrix}  + K \begin{bmatrix} e_{rot} \\p_d - {p}  \end{bmatrix} +  f_{d} \right )
     \end{align}
    The desired behaviour of \eqref{eq:impedance_desired} is achieved as the robot's acceleration approaches the target acceleration, i.e., we want to minimize the impedance task error $e_{imp}$ given by
        \begin{equation} \label{eq:e_imp_} e_{imp} = 
    \begin{bmatrix} \alpha_t \\ a_t \end{bmatrix}-  \begin{bmatrix} \alpha \\ a \end{bmatrix}
    \end{equation}
    The QP scheme requires the task to be formulated as a linear function of the optimization parameter $\ddot{q}$. The relation between the task-space and joint-space acceleration can be found using the Jacobian $J$, which adheres to
    \begin{equation}
    \begin{bmatrix} \omega \\ v \end{bmatrix}=J\dot{q}.
    \end{equation}
    Deriviating with respect to time gives
    \begin{equation} \label{eq:jacdiff}
    \begin{bmatrix} \alpha \\ a \end{bmatrix}=J\ddot{q}+\dot{J}\dot{q}.
    \end{equation}    
    Finally, combining \eqref{eq:e_imp_} and \eqref{eq:jacdiff} results in the impedance task error as a function of the optimization variable:
        \begin{equation} \label{eq:e_imp} e_{imp}(\ddot{q}) = 
    \begin{bmatrix} \alpha_t \\ a_t \end{bmatrix}-  J\ddot{q}-\dot{J}\dot{q}.
    \end{equation}
    \textbf{Posture task:} In object manipulation, the 6-DoF end effector pose is of particular interest. Robot arms that mimic human anatomy often posses 7 DoF's, however. This results in a DoF redundancy, which should be addressed for a well-posed QP scheme. The additional DoF can be imposed by a so-called posture task, as was done in \cite{vanoorschotDesignNumericalValidation2022}. This task attaches a virtual, critically-damped spring and damper with stiffness $k$ to the joint angles, i.e.,
    \begin{equation}
    \ddot{q}_t = k (q_d-q) - 2\sqrt{k}\dot{q}.
    \end{equation}
    The posture task error is then given by
            \begin{equation} \label{eq:e_pos} e_{pos}(\ddot{q}) = 
    k (q_d-q) - 2\sqrt{k}\dot{q} -\ddot{q}
    \end{equation}

    \subsection{Reference spreading}
     \textcolor{red}{goal: explain error peaking, explain reference spreading, formulize reference extending, explain intermediate mode }\\
    \textbf{Error peaking:} Many controllers, such as the impedance-based controller described in the previous section, rely on a velocity reference for tracking. When a trajectory contains impacts, both the measured and reference velocity signals contain jumps. Due to, e.g., imperfect tracking or uncertainty regarding the position of the contact surface, the measured and reference jumps can be misaligned in time. This causes the tracking error to peak: a phenomenon that will be refered to as error peaking. This is shown in Figure xxx. The peak is paired with high control efforts that could damage the robot and its environment, while not necesarily contributing to better tracking. To prevent large peaks in the tracking error, robots often avoid high-velocity impacts.

    \textbf{Switching reference:} Reference spreading (RS) enables high-velocity impacts by adressing error peaking. RS distinguishes between a reference before and after the impact. While neither of references contains velocity jumps, switching from the ante- to the post-impact reference upon impact results in a velocity jump that is aligned with the the velocity jump of the actual system. Figure xxx shows that this reduces the peak in the tracking error, meaning that faster impact become feasible.

   \textbf{Near-simultaneous impacts:} Sometimes, impacts are paired with multiple velocity jumps.  Corners of a contact surface may impact at diverging intervals in uncertain order during what is called near-simultaneous impacts. This is visualised in Figure xxx. Between the moment of impact and contact completion -- an interval refered to as the intermediate (impact) phase -- both the ante- and post impact reference would result in error peaking. Literature describes intermediate phase controller to deal with this: during the intermediate phase, velocity feedback may be disabled \cite{vansteenRobotControlSimultaneous2021}, or the velocity reference may be set to zero \cite{uitendaalTeachingRobotsInteraction2022}.

@@

% Methods
    \section{Methods}
    \subsection{Soft end effector}
    \subsection{Trajectory planning via teleoperation}
    Due to the absence of an environment model, we cannot replicate the trajectory planning methods used in simulation as described in Section xxx. Alternatively, a human could demonstrate tasks through teleoperation. The trajectory tracked by the robot during demonstration could then be used as reference as it is guaranteed to be coherent with the robot's dynamics. This section frormulates a teleoperation controller and how it can be used to plan trajectories.

@@

    \subsection{Constraining control effort}
    Section xxx explained how velocity jumps result in control effort that may damage the robot or the environment. Even though there are no impact-driven velocity jumps in a reference extracted from a non-haptic VR device, jumps in velocity measurements may still cause excessive control effort.

>>>>

% Setup
    \documentclass[a4paper, 10pt, conference]{ieeeconf}
    \IEEEoverridecommandlockouts                              % This command is only
    % needed if you want to
    % use the \thanks command

    \overrideIEEEmargins
    % See the \addtolength command later in the file to balance the column lengths
    % on the last page of the document

@@

    % The following packages can be found on http:\\www.ctan.org
    \usepackage{graphics} % for pdf, bitmapped graphics files
    \usepackage{epsfig} % for postscript graphics files
    \usepackage{mathptmx} % assumes new font selection scheme installed
    \usepackage{mathptmx} % assumes new font selection scheme installed
    \usepackage{amsmath} % assumes amsmath package installed
    \usepackage{amssymb}  % assumes amsmath package installed
\usepackage{lipsum}
    \title{\LARGE \bf
    Experimental Validation of Reference Spreading by Robotic Manipulation of Unmodeled Objects
    }

@@

    \subsection{Quadratic programing control} \label{sec:QP}
     \textcolor{red}{goal: motivate QP control, formulate Cartesian ee task, posture task, and list dynamics of motion and safety constraints }\\
     Controlling a robot for object manipulation involves, e.g., adressing redundancy and acounting for limits of the robot, while simultaneously applying a contact force and moving to a position. These objectives often clash with each other, meaning that they cannot always be perfectly satisfied. It is possible to distinguish between constraints -- control objectives that must always be satisfied -- and tasks -- objectives that should be performed optimally but not necesarily perfectly.

     Quadratic Programming (QP) relies on this philosophy. In a QP scheme, constraints and task errors are defined as a linear function of optimization parameters. It then becomes possible to efficiently find the parameters that minimize the task errors while adhering to the constraints. Furthermore, priority between the tasks can be enforced by assigning weights.

     In the context of a robot where the optimization parameters are, for example, joint accelerations $\ddot{q}$, the optimization problem can be formulated as

@@

     \textbf{Torque constraint:} To protect the robot, we can define a constraint that limits the control torque $\tau$ within a safe range, i.e.,
     \begin{equation}
       \tau_{min} \leq\tau \leq  \tau_{max} 
     \end{equation}
     This constraint must be formulated as a function of optimization parameter $\ddot q$ to be compatible with the QP scheme. The relation between $\ddot{q}$ and $\tau$ is given by the equations of motion
     \begin{equation}
        M\ddot{q}+h=\tau
     \end{equation}    
     with inertia matrix $M$ and vector of gravity, centrifugal, and Coriolis terms $h$. Subsequently, the torque constraint formulated as linear function of the optimization parameter becomes
    \begin{equation}\label{eq:eom}
     \tau_{min} \leq M\ddot{q}+h \leq  \tau_{max}.       
     \end{equation}

    \textbf{End effector task:}
    QP control allows for a trajectory to not only be defined in the joint space, but also in the Cartesian end effector space. This is useful whern the position vector $p$ and rotation matrix $R$ should be controlled. Velocity and accelerations are represented by $\omega$ and $\alpha$ for rotations, and by $v$ and $a$ for translations. We can then define an end effector task, based on the orientation and position task in \cite{cisnerosRobustHumanoidControl2018}, following

      \begin{align} \label{eq:eetask} e_{ee}(q) = 
    \begin{bmatrix} \alpha(\ddot{q}) \\ a(\ddot{q}) \end{bmatrix}
    - D \begin{bmatrix} \omega_d - {\omega} \\ v_d - v \end{bmatrix}  - K \begin{bmatrix}(\log({R}_dR^T))^{\vee }  \\p_d - {p}  \end{bmatrix}
        \end{align} with diagonal damping and stiffness matrices $D$ and $K$. The subscript $d$ indicates a desired value. Furthermore, $\alpha$ and $a$ are related to $\ddot{q}$ with the Jacobian $J$ and its derivative $\dot{J}$, i.e.,

    \begin{equation}\label{eq:jacobian}
    \begin{bmatrix} \omega \\ v \end{bmatrix}=J\dot{q},
    \quad 
    \begin{bmatrix} \alpha \\ a \end{bmatrix}=J\ddot{q}+\dot{J}\dot{q}.
    \end{equation}

    It is important to be aware of the closed-loop dynamics of the end effector task when there are unmodeled contact forces at play. For simplicity, assume a 6-DoF system where the impedance task is tracked perfectly, i.e., $e_{ee}(q)=0$. The control torque is then found using the modeled EoM \eqref{eq:eom} with $\ddot{q}$ from \eqref{eq:eetask} and \eqref{eq:jacobian}:
     \begin{align} \label{eq:mEoM} MJ^{-1}
    \left(   
D \begin{bmatrix} \omega_d - {}{\omega} \\ v_d - v \end{bmatrix}  + K \begin{bmatrix}(\log({R}_dR^T))^{\vee }  \\p_d - {p}  \end{bmatrix}
-\dot{J}\dot{q}\right)+h =\tau.
      \end{align} In reality, the EoM includes a contact wrench $\mathcal{F}$, i.e., 
     \begin{align} \label{eq:aEoM} MJ^{-1} \left(
    \begin{bmatrix} \alpha \\ a \end{bmatrix}-\dot{J}\dot{q} \right ) +h -J^T\mathcal{F}=\tau.
      \end{align}
      Substituting \eqref{eq:mEoM} into \eqref{eq:aEoM} gives the closed-loop dynamics
 \begin{align} \label{eq:ee_closeloop} 
    \begin{bmatrix} \alpha \\ a \end{bmatrix}- D \begin{bmatrix} \omega_d - {}{\omega} \\ v_d - v \end{bmatrix}  - K \begin{bmatrix}(\log({R}_dR^T))^{\vee }  \\p_d - {p}  \end{bmatrix}=JM^{-1}J^T\mathcal{F}.
      \end{align}
      Note that the rotation and position errors are related to the contact wrench through the matrix $K^{-1}JM^{-1}J^T$.

    \textbf{Posture task:} In object manipulation, the 6-DoF end effector pose is of particular interest. Robot arms that mimic human anatomy often posses 7 DoF's, however. This results in a DoF redundancy, which should be addressed for a well-posed QP scheme. The additional DoF can be imposed by a so-called posture task described in \cite{cisnerosRobustHumanoidControl2018}. This task attaches a critically-damped spring and damper with stiffness $k$ to a virtual unit mass. The results in a target joint acceleration $\beta$, i.e.,
 \begin{equation} \label{eq:beta} \beta_n = 2\sqrt{k}(\dot{q}_{n,d}-\dot{q}) +k(q_{n,d}-q_n)
    \end{equation}
    with joint index $n$. The task error is then given by
    \begin{equation} \label{eq:e_pos} e_{pos}(\ddot{q}) = 
    \ddot{q}_n - \beta_n.
    \end{equation}

    \subsection{Reference spreading}
     \textcolor{red}{goal: explain error peaking, explain reference spreading, formulize reference extending, explain intermediate mode }\\
   Many controllers rely on a velocity reference for tracking. When a trajectory contains impacts, both the measured and reference velocity signals contain jumps. Due to, e.g., imperfect tracking or uncertainty regarding the position of the contact surface, the measured and reference jumps can be misaligned in time. This causes the tracking error to peak -- a phenomenon that will be refered to as error peaking. An example of error peaking is shown in Figure xxx. The peak is paired with high control efforts that could damage the robot and its environment, while not necesarily contributing to better tracking. To prevent large peaks in the tracking error, robots often avoid high-velocity impacts.

    \textbf{Switching reference:} Reference spreading (RS) enables high-velocity impacts by adressing error peaking. RS distinguishes between a reference before and after the impact. By switching between the ante and post impact reference upon impact detection, it is possible to create a jump in the velocity reference that aligns with the velocity jump of the actual system. Figure xxx shows that this reduces the peak in the tracking error, meaning that excessive control efforts are avoided and that faster impact become feasible.

   \textbf{Near-simultaneous impacts:} Sometimes, impacts are paired with multiple velocity jumps.  Corners of a contact surface may impact at diverging intervals in uncertain order during what is called near-simultaneous impacts. This is visualised in Figure x. Similarly, in the context of dual-arm box grabbing, one arm may establish contact before the other.

   Between the moment of impact and contact completion -- an interval refered to as the interim (impact) phase -- both the ante- and post impact reference would result in error peaking. Literature describes interim phase controllers to deal with this: during the interim phase, velocity feedback may be disabled \cite{vansteenRobotControlSimultaneous2021}, or the velocity reference may be set to zero \cite{uitendaalTeachingRobotsInteraction2022}.

@@

% Methods
    \section{Methods}
    \subsection{Soft end effector}
     \begin{figure*}[h]
      \centering

    \includegraphics[trim={1cm 23.3cm 5cm 1cm}, clip]{Graphics/qp.pdf}
    \includegraphics[trim={1cm 17.8cm 5cm 6.5cm}, clip]{Graphics/qp.pdf}
      \caption{Inductance of oscillation winding on amorphous
       magnetic core versus DC bias magnetic field}
      \label{figurelabel}
   \end{figure*}

    \subsection{Trajectory planning via teleoperation}
    Due to the absence of an environment model, we cannot replicate the trajectory planning methods used in simulation as described in Section xxx. Alternatively, a human could demonstrate tasks through teleoperation. The trajectory tracked by the robot during demonstration could then be used as reference as it is guaranteed to be coherent with the robot's dynamics. This section frormulates a teleoperation controller and how it can be used to plan trajectories.

@@

     \textbf{Impedance task:} Recall that, for the end effector task \eqref{eq:eetask}, the contact wrench $\mathcal{F}$ is related to the rotation and position errors through the matrix $K^-1JM^{-1}J^T$. This is problematic in the context of intuitive teleoperation; since the operator has no sense of $JM^{-1}J^T$, performing teleoperation tasks in contact becomes practically impossible.
     An alternative can be found in impedance control -- an approach that has been proven in the context of object manipulation and teleoperation. This section describes the formulation of an impedance task.

    It is preferable to apply a virtual impedance in Cartesian space rather than enforcing a virtual impedance at each joint, as this is more intuitive when tuning task weights and gains, or when performing teleoperation. The impedance should then act on the Cartesian pose of a body, i.e., $R$ and $p$.

     An example of a virtual impedance is given by
         \begin{equation} \label{eq:impedance_desired}
     \Lambda  \begin{bmatrix} \alpha   \\ a  \end{bmatrix} + D \begin{bmatrix} \omega - {\omega_d} \\ v - v_d \end{bmatrix}  + K \begin{bmatrix} -(\log({R}_dR^T))^{\vee } \\p - {p_d}  \end{bmatrix} = 0
     \end{equation}
    with Cartesian-space inertia matrix $\Lambda$.

    Stiffness $K$ is typically chosen as a diagonal matrix. The damping matrix is determined following $D = 2(\Lambda K)^{\frac{1}{2}}$ which guarantees stable behavior when $K$ and $\Lambda$ are symmetric~\cite{albu-schafferCartesianImpedanceControl2003}.

    Based on \eqref{eq:impedance_desired}, we can define an impedance task error following
     \begin{align} \label{eq:e_imp_} e_{imp}(q) = 
    \begin{bmatrix} \alpha(\ddot{q}) \\ a(\ddot{q}) \end{bmatrix}
    - \Lambda^{-1}f 
        \end{align}
        with 
    \begin{align} f =  D \begin{bmatrix} \omega_d - {\omega} \\ v_d - v \end{bmatrix}  + K \begin{bmatrix}(\log({R}_dR^T))^{\vee }  \\p_d - {p}  \end{bmatrix}.
        \end{align}
    The closed-loop behaviour of the impedance task in a situation with unknown contact wrench $\mathcal{F}$ may be derived similiarly as was done in \eqref{eq:ee_closeloop}. This results in
     \begin{align} \label{eq:imp_closeloop} 
    \Lambda \begin{bmatrix} \alpha \\ a \end{bmatrix}- D \begin{bmatrix} \omega_d - {}{\omega} \\ v_d - v \end{bmatrix}  - K \begin{bmatrix}(\log({R}_dR^T))^{\vee }  \\p_d - {p}  \end{bmatrix}=\Lambda JM^{-1}J^T\mathcal{F}.
      \end{align}
      Note that, if $\Lambda$ were equal to identity, \eqref{eq:imp_closeloop} would be identical to \eqref{eq:ee_closeloop}. Instead, we choose $\Lambda=J^{-T}MJ^{-1}$, resulting in the closed loop dynamics 
           \begin{align} \label{eq:imp_closeloop_} 
    J^{-T}MJ^{-1}\begin{bmatrix} \alpha \\ a \end{bmatrix}- D \begin{bmatrix} \omega_d - {}{\omega} \\ v_d - v \end{bmatrix}  - K \begin{bmatrix}(\log({R}_dR^T))^{\vee }  \\p_d - {p}  \end{bmatrix}=\mathcal{F},
      \end{align}
where the contact wrench is now related to the rotation and tracking errors through the diagonal stiffness matrix -- a relation that is much more intuitive to the operator compared to the end effector task. This comes at the cost of worse tracking during free motion however, since the relation of the rotation and position error with the accelerations has become more convoluted. Nevertheless, having good control over the contact forces is prefered because since the accelerations convert to zero in steady state while the contact forces do not. (\textcolor{red}{how to improve this explanation?})

### 2023-01-07 QP.tex
aha

Note that the functions $S_{hold}$, $S_{integ}$, and $S_{rot}$ als require $\varphi_{rec}$ to be evaluated. This dependancy is excluded from notation for brevity. 
\section*{Reference Switching}

@@

\\
\begin{bmatrix} 
p_{ref}(t) \\
 q_{1,ref}(t) \\
 \dot{q}_{1,ref}(t)\\
 f_{ref}(t)\\
 \beta_{ref}(t)
\end{bmatrix} = U_{cont} \left (
\begin{bmatrix} 
p_{ext}(t,j)\\
 q_{1,ext}(t,j)\\
 \dot{q}_{1,ext}(t,j)\\
 f_{ext}(t,j)\\
 \beta_{ext}(t,j)
\end{bmatrix} 
 \right )\\
\\

\\
\begin{bmatrix} 
\omega_{ref}(t)\\
v_{ref}(t)\\
\end{bmatrix} = U_{jump} \left (

\begin{bmatrix} 
\omega_{rep}(t)\\
v_{rep}(t)\\
\end{bmatrix},

\begin{bmatrix} 
\omega_{ext}(t,j)\\
v_{ext}(t,j)\\
\end{bmatrix}

\right )\\
\\
\end{array}$
Note that the functions $U_{jump}$, $U_{cont}$, and $U_{rot}$ als require $j_{rep}$, $j_{rec}$, $\gamma$ and $m$ to be evaluated. These dependencies are excluded from notation for brevity. 
\newpage

@@

\end{document}

>>>>

Note that the functions $S_{hold}$, $S_{integ}$, and $S_{rot}$ also require $\varphi_{rec}$ to be evaluated. This dependancy is excluded from notation for brevity. 
\section*{Reference Switching}

@@

\\
\begin{bmatrix} 
p_{ref}(t) \\
 q_{1,ref}(t) \\
 f_{ref}(t)\\
 \beta_{ref}(t)
\end{bmatrix} = U_{cont} \left (
\begin{bmatrix} 
p_{ext}(t,j)\\
 f_{ext}(t,j)\\
 \beta_{ext}(t,j)
\end{bmatrix} 
 \right )\\
\\

\\
\begin{bmatrix} 
\omega_{ref}(t)\\
v_{ref}(t)\\
 \dot{q}_{1,ref}(t,j)\\
\end{bmatrix} = U_{jump} \left (

\begin{bmatrix} 
\omega_{rep}(t)\\
v_{rep}(t)\\
 \dot{q}_{1,rep}(t)\\
\end{bmatrix},

\begin{bmatrix} 
\omega_{ext}(t,j)\\
v_{ext}(t,j)\\
 \dot{q}_{1,ext}(t,j)\\
\end{bmatrix}

\right )\\
\\
\end{array}$
Note that the functions $U_{jump}$, $U_{cont}$, and $U_{rot}$ also require $j_{rep}$, $j_{rec}$, $\gamma$ and $m$ to be evaluated. These dependencies are excluded from notation for brevity. 
\newpage

@@

Need a different symbol for $\dot{q}_1$ since it is not always the derivative of $q_1$. 
\end{document}
